Length Gauge Theory of Surface Second Harmonic Generation

### 2014-11-19 paper-v-0.tex

\documentclass[floatfix,prb,aps,superscriptaddress,11pt,preprint,letterpaper]{revtex4}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\allowdisplaybreaks[1]%biutiful equation breaker!!!
\usepackage{ulem}
\usepackage{subfigure}
%\input{/Users/bms/util/definitions}
%\def\chon[1]{\color{red} #1}
\def\chon{red}
\input{definitions}
\usepackage{bm}
%%%%*****************%%%% fine hyperef 
\usepackage[backref,pdffitwindow,colorlinks,linkcolor={blue},citecolor={red}]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{showlabels}
%\usepackage{showkeys}

%%%
% notes:
% nonlocal, not non-local
% check for tense consistency
% surface SHG = SSHG
% $\bfk$-points, not $\bfk$ points
%%%

@@

Nonlinear optics has been a relevant topic in physics research for the past 
fifty years and has gained a wide range of applications in 
many areas of technology and research. In material science,
Second-harmonic generation (SHG) is of particular importance.
It was one of the first nonlinear processes to be observed and is now one
of the most used. It has proven to be very effective as an 
optical probe for material surface and interfaces. 
\cite{hughesPRB96, 
     guyot-sionnestPRB88, 
     lupkeSSR99, 
     downerPSSA01, 
     shenAPB99, 
     shenNAT89, 
     chenPRL81, 
     maytorenaPRB98}
SHG surface spectroscopy is a crucial tool for research and development
in microelectronics, semiconductors, nanomaterials, and many more recent 
areas of great scientific and commercial interest.

Numerous mathematical frameworks have been developed to study the 
nonlinear optical response of materials; 
in the context of this paper, we note two in particular. 
The first, called the ``velocity gauge'' or 
``$\mathbf{v}\cdot\mathbf{A}$ gauge,'' features in early 
research in this field. 
Butcher and McLean\cite{butcherPOPS63}, and 
Aspnes\cite{aspnesPRB72} 
laid the initial groundwork by deriving explicit expressions for the 
nonlinear susceptibility tensor. However, these either diverged at 
low frequencies or were limited to zincblende crystal structures. 
A few decades later, 
Ghahramani \emph{et al}.\cite{ghahramaniPRB91} 
realized a full-band-structure calculation for Si/Ge superlattices and 
found significant surface SHG while providing proof that the expressions 
do not diverge at low frequencies for clean, cold semiconductors. 
Sipe \emph{et al}.\cite{sipePRB93}
separated the interband and intraband one-electron contributions 
of the susceptibility tensor and further verified the non-divergent behavior. 
Levine\cite{levinePRB94} 
introduced the scissors operator within the Local Density Approximation
(LDA) and found good agreement with experimental results. 
The second, called the ``length gauge'' or 
``$\mathbf{r}\cdot\mathbf{E}$ gauge'' has various advantages over 
the velocity gauge. 
Aversa and Sipe\cite{aversaPRB95} 
presented divergence free expressions for the susceptibility tensor using 
the length gauge that are significantly easier to derive and calculate. 
Cabellos \emph{et al}.\cite{cabellosPRB09} 
elaborated a detailed derivation of the length gauge 
formalism and a comparison with the velocity gauge for bulk GaAs. They 
proved that the results are indeed gauge-invariant. 
This presents a strong case for using the length over the velocity gauge 
for more generalized and less expensive calculations of the nonlinear
response of a system.

The inclusion of nonlocal effects, such as the scissors correction, has been
correctly included in both the velocity and length guage schemes. 
In the former, both 
Rashkeev \emph{et al}.\cite{rashkeevPRB98} and 
Hughes and Sipe\cite{hughesPRB96} 
accounted for the scissors correction when calculating the nonlinear 
susceptibility tensor.
Leitsmann \emph{et al}.\cite{leitsmannPRB05} 
included excitonic and local field interactions working with GaAs.
Nastos \emph{et al}.\cite{nastosPRB05} demonstrated the correct 
way to calculate the nonlinear response using the scissored Hamiltonian 
in the length gauge providing better agreement with experiment.

Concerning the optical response of surfaces and interfaces,
Mendoza and Moch\'an\cite{mendozaPRB96a} produced an early review of 
theoretical SHG spectra from surfaces. 
Arzate and Mendoza\cite{arzatePRB01} improved upon that work and
went into a detailed analysis of the different contributions to the 
SHG spectra of a surface and the nuanced relationship between bulk, 
surface, interband, intraband, $1\omega$ and $2\omega$ terms. 
Calculated surface SHG spectra using these schemes coincide with experimental
results in both the velocity gauge
\cite{mendozaPRB01,mendozaPRL98,reiningPRB94} and the length gauge approach
\cite{sharmaPRB03}. 
Perhaps of more relevance to this work, 
Mendoza \emph{et al}.\cite{mendozaPRB06} and 
Mej\'ia \emph{et al}.\cite{mejiaRMF04} developed a layer-by-layer analysis for 
the linear and nonlinear responses of semiconductor systems. This model
allows for obtaining results from selected regions within a slab or bulk,
including the surface.
More recently, 
Trolle \emph{et al}.\cite{trollePRB14} 
elaborate on a new theoretical model for surface SHG of 
mono-layer MoS$_{2}$. Their approach is based on the tight-binding model 
fitted to GW results and include excitonic effects from the 
Bethe-Salpeter exciton picture. However, this new model is not 
\emph{ab initio} and has no consideration for a layered approach.

The most complete approach will be a combination of the different techniques 
that lead to more efficient and accurate calculations. Therefore,
the goal of this paper is to present newly derived expressions for the 
nonlinear susceptibility tensor $\chi^{abc}(-2\omega;\omega,\omega)$ 
in the length gauge formalism
that include nonlocal terms from the scissors approximation 
and the spatial contribution from the pseudopotentials and a 
layer-by-layer approach convenient for surface SHG calculations. We are
able to divide the resulting tensor in interband, intraband, $1\omega$, and 
$2\omega$ contributions. This approach is unprecedented and yet produces 
results that are in complete agreement with the literature.

The paper is organized as follows. 
In Sec. \ref{theory} we present the 
relevant steps for deriving the surface second-order susceptibility tensor 
within the length gauge formalism. This derivation includes the addition of 
the nonlocal terms mentioned above and a layer-by-layer analysis that have 
been absent in previous works. 
In Sec. \ref{results} we show the results of this 
reformulation with a layer-by-layer study of a clean Si(111) surface with a 
$2 \times 1$ reconstruction. These calculations are done with \emph{ab initio}
programs based on density-functional theory (DFT) within the local-density 
approximation (LDA) using Troullier-Martins pseudopotentials. We compare 
results from before and after adding the different nonlocal contributions.
Finally, in Sec. \ref{conclusions} we submit our conclusions.

@@

We use the independent particle approximation and neglect local field
and excitonic effects{\color{\chon}. We assume a classical} electromagnetic field
{\color{\chon} and quantum mechanical matter.}
We can describe the system using 
%a scaled 
the one electron density
operator ${\rho}$, with which we can calculate the expectation value of a
single-particle observable $\mathcal{O}$ as 
$\langle{\cal O}\rangle=\mbox{Tr}({\rho} {\cal O}),  
\label{calo}$
with $\mathcal{O}$ the associated quantum mechanical operator and Tr
the trace. 
The density operator satisfies
$
i\hbar (d{\rho}/dt)=[H(t),{\rho}],  
\label{rho}
$
with $H(t)$ as the total single electron Hamiltonian, written as 
\begin{equation*}
H(t)=H_{0}+H_{I}(t),  
\label{ache}
\end{equation*}
where $H_{0}$ is the unperturbed time-independent Hamiltonian, and $H_{I}(t)$
is the time-dependent potential energy due to the interaction of the
electron with the electromagnetic field.
% ; $H_{0}$ has eigenvalues
% $\hbar \omega_{n}(\mathbf{k})$
% and eigenstates $| n\mathbf{k} \rangle$ (Bloch states) labeled by a band 
% index $n$ and crystal momentum $\mathbf{k}$.
To proceed with the solution of $\rho$  it is convenient to use the
interaction picture, where a unitary operator $U=\exp ({iH_{0}t/\hbar })$
transforms any operator $\mathcal{O}$ into 
$\tilde{\mathcal{O}}=U{\cal O}U^{\dagger }$. Even if $\mathcal{O}$ 
{\color{\chon} is time-independent}, 
$\tilde{\mathcal{O}}$ {\color{\chon} is time-dependent} through the explicit
time dependence of $U$. 
The dynamical 
equation for $\tilde{\rho}$ is 
given by
\begin{equation*}
i\hbar \frac{d\tilde{{\rho}}}{dt}=[\tilde{H}_{I}(t),\tilde{{\rho}}],  
\label{rho1}
\end{equation*}
with solution 
\begin{equation}
i\hbar \tilde{{\rho}}(t)=i\hbar \tilde{{\rho}}_{0}+\int_{-\infty }^{t}dt^{\prime }[
\tilde{H}_{I}(t^{\prime }),\tilde{\rho}(t^{\prime })],  
\label{trans}
\end{equation}
where $\tilde{\rho}_{0}=\tilde{\rho}(t=-\infty )$ is the unperturbed density matrix. We look
for the standard perturbation series solution, 
$\tilde{\rho}(t)=\tilde{\rho}^{(0)}+\tilde{\rho}^{(1)}+\tilde{\rho}^{(2)}+\cdots$,  
where the superscript denotes the order (power) with which each term depends
on the perturbation $H_{I}(t)$. From Eq.~\eqref{trans} the $N$-th order
term is 
\begin{equation}
\tilde{{\rho}}^{(N)}(t)=\frac{1}{i\hbar }\int_{-\infty }^{t}dt^{\prime }[\tilde{
H}_{I}(t^{\prime }),\tilde{\rho}^{(N-1)}(t^{\prime })].  
\label{rhop}
\end{equation}
The series is generated by the unperturbed density operator $\tilde{\rho}
^{(0)}\equiv \tilde{\rho}_{0}$, assumed to be the diagonal Fermi-Dirac distribution, 
$\langle n\mathbf{k}|\tilde{\rho}_{0}|n\mathbf{k}\rangle=f(\hbar \omega_{n}(\mathbf{k}))\equiv f_{n}$. For a
clean, cold semiconductor $f_{n}=1$ {\color{\chon} when} $n$ {\color{\chon} is} a valence ($v$) or
occupied band, and zero {\color{\chon} when} $n$ {\color{\chon} is} a conduction ($c$) or empty band.
{\color{\chon} We assume this for the remainder of the article.}
{\color{\chon} T}he expectation values must satisfy
$
\langle{\cal O}\rangle=
\mbox{Tr}({\rho}{\cal O})
=
\mbox{Tr}(\tilde{{\rho}}\tilde{\cal O})
$.

We will look for the expectation value of the macroscopic current density, 
$\mathbf{J}$, given by 
\begin{equation}
\bfJ=\langle{\mathbf{J}}\rangle=\frac{e}{\Omega}\mbox{Tr}({\rho}\dot{\mathbf{r}})
,
\label{pe}
\end{equation}
where $\dot{\mathbf{r}}$ is the time derivative of the position operator of the
electron {\color{\chon} with} charge $e$, {\color{\chon} defined as}
\begin{equation}
\mathbf{v}\equiv \dot{\mathbf{r}}=\frac{1}{i\hbar }[\mathbf{r},H_0],  
\label{mv}
\end{equation}
with $\mathbf{v}$ the velocity operator of the electron, and $\Omega$ the
normalization volume. We calculate the macroscopic polarization density 
$\mathbf{P}$, related to $\bfJ$ by
$\bfJ=d\mathbf{P}/dt$. For a 
perturbing (Maxwell macroscopic) electromagnetic field, $\mathbf{E}(t)=
\mathbf{E}(\omega )e^{-i\tilde{\omega} t}+c.c.$,
where $\tilde\omega=\omega+i\eta $,
and $\eta >0$ {\color{\chon} adiabatically switches on} the interaction,
we write the second order nonlinear
polarization as, 
\begin{equation}
P^a(2\omega)=\chi ^{abc}(-2\omega;\omega,\omega)E^{b}(\omega)E^{c}(\omega),  
\label{pshg}
\end{equation}
where $\chi^{abc}(-2\omega ;\omega ,\omega )$ is the nonlinear
susceptibility responsible {\color{\chon} for} surface second harmonic generation
(SSHG). 
The 
superscripts in Eq.~\eqref{pshg} denote Cartesian components, and if
repeated are to be summed over. Without loss of generality we can always
define $\chi^{abc}(-2\omega;\omega,\omega)$
 to satisfy intrinsic permutation
symmetry, 
$\chi^{abc}(-2\omega ;\omega ,\omega )=\chi ^{acb}(-2\omega ;\omega
,\omega )$.
% ;
% if it did not, the part that did not satisfy intrinsic permutation symmetric
% would make no contribution to the second order polarization. This part
% could be dropped since it would
% have no physical significance.

The unperturbed Hamiltonian 
is used to solve the Kohn-Sham equations\cite{kohnPR65} of Density  
Functional Theory (DFT) {\color{\chon}. It is convenient to work within the Local 
Density Approximation (LDA), so we label the hamiltonian with the corresponding  
LDA superscript. Any other approximation can be used (like GGA) and our 
derivation remains the same.} Then,
\begin{equation}\label{h0}
H^{\mathrm{LDA}}_{0}=\frac{p^{2}}{2m_e}+V
\end{equation}
with $m_e$ the mass of the electron, $\mathbf{p}$ its canonical momentum, and 
$V$
 the periodic crystal potential, where we neglect spin-orbit terms.
To be more general in our derivation of
$\chi^{abc}(-2\omega;\omega,\omega)$, we asume
that the crystal potential could have a {\color{\chon} non-local spatial}
contribution {\color{\chon} as} is customary for most
pseudopotentials, and then we replace $V$ with
\begin{align}\label{ps}
V^{\mathrm{ps}}=V(\bfr)+V^{\mathrm{nl}}
,
\end{align}
where 
$V(\bfr)$ and $V^{\mathrm{nl}}$  are the local and 
non-local contributions to the pseudopotential $V^{\mathrm{ps}}$,
 respectively.
In case of
a local potential, i.e. $V= V(\bfr)$, 
like that of all-electron schemes, 
we simply omit the contribution coming
from the non-local part of the pseudopotential, $V^{\mathrm{nl}}$, in
the results that we {\color{\chon} have derived.}

{\color{\chon} It} is well known that the use of the {\color{\chon} LDA} 
leads to an underestimation of the band gap. A standard
procedure to correct for this is to
use {\color{\chon} the} ``scissors approximation'', {\color{\chon} where the 
conduction bands are rigidly shifted in energy} so that the band gap 
corresponds to the accepted experimental band gap{\color{\chon} .
This} is often in fairly good agreement with the GW
band gap based on a more sophisticated calculation.\cite{hybertsenPRB86}
{\color{\chon} T}he LDA wave functions are used since they produce band
structures with dispersion relations similar to those predicted by the GW
approximation. Mathematically, {\color{\chon} the scissors (non-local) operator 
$S$ is added} to the unperturbed or unscissored Hamiltonian {\color{\chon} $H^{\mathrm{LDA}}_{0}$ ,}
\begin{equation*}
H^\sigma_{0}=H^{\mathrm{LDA}}_{0}+S
\end{equation*}
where 
\begin{equation}
S=\hbar \sigma\sum_{n}\int d^{3}k(1-f_{n})
|n\mathbf{k}\rangle\langle n\mathbf{k}|,
\label{hats}
\end{equation}
with $\hbar \sigma$  the rigid ($\mathbf{k}$-independent) energy correction to be
applied. 
%Several properties of $S(\mathbf{r},\mathbf{p})$ are shown in the
%appendix.
The unscissored and scissored Hamiltonians satisfy 
\begin{eqnarray*}
H^{\mathrm{LDA}}_{0}\psi _{n\mathbf{k}}(\mathbf{r}) &=&\hbar \omega^{\mathrm{LDA}}_{n}(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),
\label{hamils} \\
H_{0}^\sigma\psi _{n\mathbf{k}}(\mathbf{r}) &=&\hbar \omega_{n}^\sigma
(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),
\end{eqnarray*}
where the scissor-shifted energies, 
$\omega_{n}^\sigma(\mathbf{k})$ , are given by
\begin{equation}
\omega_{n}^\sigma(\mathbf{k})=\omega^\lda_{n}(\mathbf{k})+(1-f_{n})\sigma.
\label{wese}
\end{equation}
We emphasize that the {\color{\chon} scissored and unscissored Hamiltonian 
have the same eigenfunctions,} where
$\psi_{n\bfk}(\bfr)=\braket{\bfr}{n\bfk}=e^{i\bfk\cdot\bfr}u_{n\bfk}(\bfr)$,
are the real space representations of the Bloch states $\ket{n\bfk}$ labelled 
by the band index $n$ and the crystal momentum $\bfk$, and $u_{n\bfk}(\bfr)$
is cell periodic.

@@

cut function $\calc^\ell(z)$ in
the calculation of $\chi_{\rma\rmb\rmc}$, we start from 

@@

the entire slab gives the averaged macroscopic current density, $\bfJ(t)$. 

@@

$\ell$-th layer of the slab is given by

@@

where $\mathbf{J}^{\ell,(N)}(t)$ is the current of the

>>>>

\documentclass[floatfix,prb,aps,superscriptaddress,11pt,preprint,letterpaper]{revtex4}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\allowdisplaybreaks[1]%biutiful equation breaker!!!
\usepackage{ulem}
\usepackage{subfigure}
%\input{/Users/bms/util/definitions}
%\def\chon[1]{\color{red} #1}
\def\chon{black}
\def\chonn{red}
\def\nicolas{red}
\input{definitions}
\usepackage{bm}
%%%%*****************%%%% fine hyperef 
\usepackage[backref,pdffitwindow,colorlinks,linkcolor={blue},citecolor={red}]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{showlabels}
%\usepackage{showkeys}

%%%
% notes:
% nonlocal, not non-local
% check for tense consistency
% surface SHG = SSHG
% $\bfk$-points, not $\bfk$ points
% Nature Materials de valerie
% Local fields?
%%%

@@

In recent years, surface nonlinear optical spectroscopies and in 
particular, surface second-harmonic generation SSHG have evolved as useful 
nondestructive and noninvasive tool to study properties of surfaces 
and interfaces: atomic structure, phase transitions, and adsorption of 
atoms,1–11{\color{\chonn} [incluye estas de mejiaPRB02 y combinalas con
  las tuyas y tambien pon la de salazarPRB14 y busca referencia para
  ??, checando q sean relevantes con SSHG]} 
among others. \cite{hughesPRB96, 
     guyot-sionnestPRB88, 
     lupkeSSR99, 
     downerPSSA01, 
     shenAPB99, 
     shenNAT89, 
     chenPRL81, 
    mendozaPRL98,downerSIA01,cazzanelliNM14}
Nowadays, SSHG spectroscopy is a crucial tool for research and development
in microelectronics,\cite{??} semiconductors,\cite{??} nanomaterials,\cite{??} and many more recent 
areas of great scientific and commercial interest.\cite{??}
The high surface sensitivity of SSHG spectroscopy is due to the
fact that, within the dipole approximation, the bulk SHG signal of
centrosymmetric materials is identically zero, and thus only the
surface, where the inversion symmetry is broken, can radiate. 
Much of the foundation of surface science has been built from
experiments involving emission or scattering of electrons from
surfaces. These require ultrahigh vacuum (UHV) environments and
provide no access to buried interfaces. 
However, SSHG
is compatibile with non-UHV ambients and has access to
interfaces buried beneath transparent overlayers. Even when applied to
surfaces in UHV, the light source and detectors can be aligned and
used outside the vacuum chamber.

The usefulness of SSHG could be limited by
the lack of microscopic theoretical
understanding of the nonlinear spectra. The macroscopic
phenomenological theory of SSHG, which relates the intensity,
phase and polarization of detected fields to nonlinear and linear
susceptibilities at the material interface, is now fully developed. 
However, microscopic theory, which relates atomic-level structure to
the nonlinear source polarization, is still being developed.
In this article we put forward an approach to calculate the
microscopic second harmonic surface susceptibility that encompasses several
theoretical features not taken into account before. 
Within the independent particle approximation (IPA),  
some frameworks for bulk SHG have been developed to study the 
nonlinear optical response of bulk materials, where several degrees of
sophistication were 
introduced.\cite{butcherPOPS63,aspnesPRB72,sipePRB93,levinePRB94,aversaPRB95,hughesPRB96,rashkeevPRB98}

As it is well known
Density Functional Theory (DFT) within the Local Density Approximation
(LDA),\cite{??} the most 
used framework for \textit{ab initio} calculations,
underestimates the energy band gap of semiconductors. It is also well
understood that one has to include the many-body interaction to
correct for the underestimation of the gap. In this context the so
called GW approximation\cite{onidaRMP01} is known to correct the band
gap of most semiconductors\cite{cual}. However this could be a numerically very
expensive calculation, and thus one uses the much simpler scissors
operator scheme.\cite{?} This allows to ``open'' the DFT-LDA gap to
its correct experimental or GW value, when experiments are not available. 
In this context,
to correct for the underestimation of the energy band gap of semiconductors, 
Nastos et al.\cite{nastosPRB05} 
using the 
``length gauge'' or 
``$\mathbf{r}\cdot\mathbf{E}$ gauge'' 
showed how to correctly include in the SH susceptibility the many-body corrections through the
scissors operator.
Later,
Cabellos \textit{et al}.\cite{cabellosPRB09} 
elaborated a derivation of the ``velocity gauge'' or
``$\bfA\cdot\bfv$'' gauge properly including the scissors
operator, and proved  gauge 
invariance with respect to the length gauge.
From these works, it is clear the length gauge is a much better staring
point to obtain the surface SH (SSH) susceptibility, as it will be
elaborated in this article. Above only being valid for bulk semiconductors.

Concerning the optical response of surfaces and interfaces,
Reining \textit{et al.}\cite{reiningPRB94}  
introduced the concept of a 
cut function in order to obtain the surface SH 
susceptibility tensor.
This cut function is required since one usually uses a slab approach
when treating semi-infinite surface systems; if the slab is
centrosymmetric the susceptibility tensor will be identically zero.
The cut function is such that it separates the nonlinear response
 the two surfaces of the slab, thus avoiding the destructive
 interference between them, giving a finite value of the tensor, that
 one identifies with the SSH susceptibility tensor; if the slab is not
 centrosymmetric the cut function can be used to separate the
 different signals coming from either surface of the slab.
Indeed, one of the main results of this article is to show that the SSH 
susceptibility tensor obtained by using the cut function is 
correctly extracted from the slab.   
After Reining \textit{et al.}\cite{reiningPRB94}, 
Refs.~\onlinecite{mendozaPRL98,arzatePRB01,mendozaPRB01,mejiaPRB02}
followed upon this work and in particular Ref.~\onlinecite{arzatePRB01}
went into a detailed analysis of the different contributions to the 
SHG spectra of a surface and the nuanced relationship between bulk, 
surface, interband, intraband, $1\omega$ and $2\omega$ terms. 
Perhaps of more relevance to this work, 
Mej\'ia \textit{et al}.\cite{mejiaRMF04} developed a layer-by-layer analysis for 
the nonlinear responses of semiconductor systems, within a 
Tight-Binding framework. This model 
allows for obtaining results from selected regions of a system,
including the surface. 
However, in these references the scissors is either not implemented or
wrongly implemented.

Finally,  DFT-LDA calculations are often based on the use of pseudopotentials
that usually contain spatial non-localities to include the interaction
of the valence electrons with the core electrons.
As it will be discussed in this article, 
 this non-local
dependence on the position vector, introduces corrections to
the velocity operator of the electron, that have to be included with
care in the SSH susceptibility. So far, this contribution has been
neglected in the literature.

Therefore, within the IPA, 
the most complete approach 
for the calculation of the SSH susceptibility is one which includes
i) the scissors correction, ii) the cut function, and iii) the  
contribution of the non-local part of the pseudopotentials.  
Therefore,
the goal of this paper is to derive, within the length gauge, 
new expressions for the 
SSH susceptibility tensor 
$\chi^{\rma\rmb\rmc}(-2\omega;\omega,\omega)$ 
that include the aforementioned contributions.
The inclusion of these three contributions 
makes 
our scheme unprecedented and opens the possibility to study 
surface SHG with more versatility and providing 
more accurate results than previous approaches.

The paper is organized as follows. 
In Sec. \ref{theory} we present the 
relevant steps for deriving the surface second-order susceptibility
tensor
$\chi^{\rma\rmb\rmc}(-2\omega;\omega,\omega)$  
within the length gauge formalism. This derivation includes the addition of 
the terms mentioned above
that have 
been absent in previous works. Also,
our $\chi^{\rma\rmb\rmc}(-2\omega;\omega,\omega)$  
can be used for a layer-by-layer
analysis if wanted.
In Sec. \ref{results} we show the results of this 
reformulation with a study of a clean Si(001) surface with a 
$2 \times 1$ reconstruction. 
We compare 
results from before and after adding the different nonlocal contributions.
Finally, in Sec. \ref{conc} we give our conclusions.

@@

We use the IPA and neglect local field  
and excitonic effects{\color{\chon}.  
We assume a classical} electromagnetic field
{\color{\chon} and quantum mechanical matter.}
We can describe the system using 
%a scaled 
the one electron density
operator ${\rho}$, with which we can calculate the expectation value of a
single-particle observable $\mathcal{O}$ as 
$\langle{\cal O}\rangle=\mbox{Tr}({\rho} {\cal O}),  
\label{calo}$
with $\mathcal{O}$ the associated quantum mechanical operator and Tr
the trace. 
The density operator satisfies
$
i\hbar (d{\rho}/dt)=[H(t),{\rho}],  
\label{rho}
$
with $H(t)$ as the total single electron Hamiltonian, written as 
\begin{equation*}
H(t)=H_{0}+H_{I}(t),  
\label{ache}
\end{equation*}
where $H_{0}$ is the unperturbed time-independent Hamiltonian, and $H_{I}(t)$
is the time-dependent potential energy due to the interaction of the
electron with the electromagnetic field.
% ; $H_{0}$ has eigenvalues
% $\hbar \omega_{n}(\mathbf{k})$
% and eigenstates $| n\mathbf{k} \rangle$ (Bloch states) labeled by a band 
% index $n$ and crystal momentum $\mathbf{k}$.
To proceed with the solution of $\rho$  it is convenient to use the
interaction picture, where a unitary operator $U=\exp ({iH_{0}t/\hbar })$
transforms any operator $\mathcal{O}$ into 
$\tilde{\mathcal{O}}=U{\cal O}U^{\dagger }$. Even if $\mathcal{O}$ 
{\color{\chon} is time-independent}, 
$\tilde{\mathcal{O}}$ {\color{\chon} is time-dependent} through the explicit
time dependence of $U$. 
The dynamical 
equation for $\tilde{\rho}$ is 
given by
\begin{equation*}
i\hbar \frac{d\tilde{{\rho}}}{dt}=[\tilde{H}_{I}(t),\tilde{{\rho}}],  
\label{rho1}
\end{equation*}
with solution 
\begin{equation}
i\hbar \tilde{{\rho}}(t)=i\hbar \tilde{{\rho}}_{0}+\int_{-\infty }^{t}dt^{\prime }[
\tilde{H}_{I}(t^{\prime }),\tilde{\rho}(t^{\prime })],  
\label{trans}
\end{equation}
where $\tilde{\rho}_{0}=\tilde{\rho}(t=-\infty )$ is the unperturbed density matrix. We look
for the standard perturbation series solution, 
$\tilde{\rho}(t)=\tilde{\rho}^{(0)}+\tilde{\rho}^{(1)}+\tilde{\rho}^{(2)}+\cdots$,  
where the superscript denotes the order (power) with which each term depends
on the perturbation $H_{I}(t)$. From Eq.~\eqref{trans} the $N$-th order
term is 
\begin{equation}
\tilde{{\rho}}^{(N)}(t)=\frac{1}{i\hbar }\int_{-\infty }^{t}dt^{\prime }[\tilde{
H}_{I}(t^{\prime }),\tilde{\rho}^{(N-1)}(t^{\prime })].  
\label{rhop}
\end{equation}
The series is generated by the unperturbed density operator $\tilde{\rho}
^{(0)}\equiv \tilde{\rho}_{0}$, assumed to be the diagonal Fermi-Dirac distribution, 
$\langle n\mathbf{k}|\tilde{\rho}_{0}|n\mathbf{k}\rangle=f(\hbar \omega_{n}(\mathbf{k}))\equiv f_{n}$. For a
clean, cold semiconductor $f_{n}=1$ {\color{\chon} when} $n$ {\color{\chon} is} a valence ($v$) or
occupied band, and zero {\color{\chon} when} $n$ {\color{\chon} is} a conduction ($c$) or empty band.
{\color{\chon} We assume this for the remainder of the article.}
{\color{\chon} T}he expectation values must satisfy
$
\langle{\cal O}\rangle=
\mbox{Tr}({\rho}{\cal O})
=
\mbox{Tr}(\tilde{{\rho}}\tilde{\cal O})
$.

We will look for the expectation value of the microscopic current density, 
$\mathbf{J}$, given by 
\begin{equation}
\bfJ=\langle{\mathbf{J}}\rangle=\frac{e}{A}\mbox{Tr}({\rho}\dot{\mathbf{r}})
,
\label{pe}
\end{equation}
where $\dot{\mathbf{r}}$ is the time derivative of the position operator of the
electron {\color{\chon} with} charge $e$, {\color{\chon} defined as}
\begin{equation}
\mathbf{v}\equiv \dot{\mathbf{r}}=\frac{1}{i\hbar }[\mathbf{r},H_0],  
\label{mv}
\end{equation}
with $\mathbf{v}$ the velocity operator of the electron, and $A$ the
area of the unit cell. We calculate the macroscopic polarization density 
$\mathbf{P}$, related to $\bfJ$ by
$\bfJ=d\mathbf{P}/dt$. For a 
perturbing (Maxwell macroscopic) electromagnetic field, $\mathbf{E}(t)=
\mathbf{E}(\omega )e^{-i\tilde{\omega} t}+c.c.$,
where $\tilde\omega=\omega+i\eta $,
and $\eta >0$ {\color{\chon} adiabatically switches on} the interaction,
we write the second order nonlinear
polarization as, 
\begin{equation}
\calp^a(2\omega)=\chi ^{abc}(-2\omega;\omega,\omega)E^{b}(\omega)E^{c}(\omega),  
\label{pshg}
\end{equation}
where $\chi^{abc}(-2\omega ;\omega ,\omega )$ is the nonlinear
susceptibility responsible {\color{\chon} for} surface second harmonic generation
(SSHG). 
The 
superscripts in Eq.~\eqref{pshg} denote Cartesian components, and if
repeated are to be summed over. Without loss of generality we can always
define $\chi^{abc}(-2\omega;\omega,\omega)$
 to satisfy intrinsic permutation
symmetry, 
$\chi^{abc}(-2\omega ;\omega ,\omega )=\chi ^{acb}(-2\omega ;\omega
,\omega )$.
% ;
% if it did not, the part that did not satisfy intrinsic permutation symmetric
% would make no contribution to the second order polarization. This part
% could be dropped since it would
% have no physical significance.

The unperturbed Hamiltonian 
is used to solve the Kohn-Sham equations\cite{kohnPR65} of Density  
Functional Theory (DFT) {\color{\chon}. It is convenient to work within the Local 
Density Approximation (LDA), so we label the hamiltonian with the corresponding  
LDA superscript. Any other approximation can be used (like GGA) and our 
derivation remains the same.} Then,
\begin{equation}\label{h0}
H^{\mathrm{LDA}}_{0}(\bfr,\bfp)
=\frac{p^{2}}{2m_e}+V(\bfr,\bfp)
\end{equation}
with $m_e$ the mass of the electron, $\mathbf{p}$ its canonical momentum, and 
$V$
 the periodic crystal potential, where we neglect spin-orbit terms.
To be more general in our derivation of
$\chi^{abc}(-2\omega;\omega,\omega)$, we asume
that the crystal potential could have a {\color{\chon} non-local spatial}
contribution {\color{\chon} as} is customary for most
pseudopotentials, and then we replace $V$ with
\begin{align}\label{ps}
V^{\mathrm{ps}}(\bfr,\bfp)=V(\bfr)+V^{\mathrm{nl}}(\bfr,\bfp)
,
\end{align}
where 
$V(\bfr)$ and $V^{\mathrm{nl}}(\bfr,\bfp)$  are the local and 
non-local contributions to the pseudopotential $V^{\mathrm{ps}}$,
 respectively.
In case of
a local potential, i.e. $V= V(\bfr)$, 
like that of all-electron schemes, 
we simply omit the contribution coming
from the non-local part of the pseudopotential, $V^{\mathrm{nl}}$, in
the results that we {\color{\chon} have derived.}

{\color{\chon} It} is well known that the use of the {\color{\chon} LDA} 
leads to an underestimation of the band gap. A standard
procedure to correct for this is to
use {\color{\chon} the} ``scissors approximation'', {\color{\chon} where the 
conduction bands are rigidly shifted in energy} so that the band gap 
corresponds to the accepted experimental band gap{\color{\chon} .
This} is often in fairly good agreement with the GW
band gap based on a more sophisticated calculation.\cite{hybertsenPRB86}
{\color{\chon} T}he LDA wave functions are used since they produce band
structures with dispersion relations similar to those predicted by the GW
approximation. Mathematically, {\color{\chon} the scissors (non-local) operator 
$S$ is added} to the unperturbed or unscissored Hamiltonian {\color{\chon} $H^{\mathrm{LDA}}_{0}$ ,}
\begin{equation*}
H^\sigma_{0}(\bfr,\bfp)=H^{\mathrm{LDA}}_{0}(\bfr,\bfp)+S(\bfr,\bfp)
\end{equation*}
where 
\begin{equation}
S(\bfr,\bfp)=\hbar \sigma\sum_{n}\int d^{3}k(1-f_{n})
|n\mathbf{k}\rangle\langle n\mathbf{k}|,
\label{hats}
\end{equation}
with $\hbar \sigma$  the rigid ($\mathbf{k}$-independent) energy correction to be
applied. 
%Several properties of $S(\mathbf{r},\mathbf{p})$ are shown in the
%appendix.
The unscissored and scissored Hamiltonians satisfy 
\begin{eqnarray*}
H^{\mathrm{LDA}}_{0}(\bfr,\bfp)\psi _{n\mathbf{k}}(\mathbf{r}) &=&\hbar \omega^{\mathrm{LDA}}_{n}(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),
\label{hamils} \\
H_{0}^\sigma (\bfr,\bfp)\psi _{n\mathbf{k}}(\mathbf{r}) &=&\hbar \omega_{n}^\sigma
(\mathbf{k})\psi _{n\mathbf{k}}(\mathbf{r}),
\end{eqnarray*}
where the scissor-shifted energies, 
$\omega_{n}^\sigma(\mathbf{k})$ , are given by
\begin{equation}
\omega_{n}^\sigma(\mathbf{k})=\omega^\lda_{n}(\mathbf{k})+(1-f_{n})\sigma.
\label{wese}
\end{equation}
We emphasize that the {\color{\chon} scissored and unscissored Hamiltonian 
have the same eigenfunctions,} where
$\psi_{n\bfk}(\bfr)=\braket{\bfr}{n\bfk}=e^{i\bfk\cdot\bfr}u_{n\bfk}(\bfr)$,
are the real space representations of the Bloch states $\ket{n\bfk}$ labelled 
by the band index $n$ and the crystal momentum $\bfk$, and $u_{n\bfk}(\bfr)$
is cell periodic.

@@

cut function $\calc(z)$ in
the calculation of $\chi^{\rma\rmb\rmc}$, we start from 

@@

the entire slab gives the averaged microscopic current density, $\bfJ(t)$. 

@@

chosen layer of the slab is given by

@@

where $\calbj^{(N)}(t)$ is the current of the
